Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-20 paper.tex
Narrative updates for additional benchmarks.

The data shown in Fig~\ref{fig-whole-matrix-compute} for Zarr and Savvy
is based on custom programs written using their respective APIs
to implement the \texttt{af-dist} operation. The Zarr program uses
the Zarr-Python package to iterate over the decoded chunks of the 
genotype matrix and classifies genotypes within a chunk using a 14 line Python
function, accelerated using the Numba JIT compiler~\cite{lam2015numba}.
The allele frequencies and genotype counts are then analysed to produce 
the final counts within the allele frequency bins with 9 lines of 
Python using NumPy~\cite{harris2020array} functions. Remarkably, this 
short and simple Python program is substantially faster than the 
equivalent compiled C using \texttt{htslib} APIs on BCF (6.9 hours
vs 20.6 hours for 1 million samples). 
% num_samples  num_sites      tool  user_time  sys_time     wall_time  total_time
%     1000000    7254858     savvy    8371.07      4.16   8377.718136  2.326453
%     1000000    7254858  bcftools   73939.32     46.09  74023.083678  20.551503
%     1000000    7254858      zarr   24709.64     29.32  24750.704171  6.871933
The fastest method is the 
C++ program written using the Savvy API. This would largely seem
to be due to Savvy's excellent genotype decoding performance
% zarr 1.2 GiB
% savvy 6.6 GiB
% zarr_nshf 3.9 GiB
(up to 6.6GiB/s vs 1.2GiB/s for Zarr on this dataset;
Fig~\ref{fig-whole-matrix-decode}).
% sequence_length  num_samples  num_sites       tool       size
% 40         48129895      1000000    7254858       zarr  22.071687
% 35         48129895      1000000    7254858        bcf  51.749294
% 36         48129895      1000000    7254858    genozip  10.691384
% 37         48129895      1000000    7254858        sav  21.249436
% 38         48129895      1000000    7254858        tsk   1.802636
% 39         48129895      1000000    7254858        vcf  81.375831
% 41         48129895      1000000    7254858  zarr_nshf  29.897540
Turning off the BitShuffle filter for the Zarr dataset,
however, leads to a substantial increase in decoding speed
(3.9GiB/s) at the cost of a roughly 25\% increase in storage
space (29.9GiB up from 22.1GiB for 1 million samples; data not
shown). The LZ4 codec (again without BitShuffle) leads to slightly higher
decoding speed, but at the cost of much larger storage requirements 
(219GiB; data not shown).
Given the relatively small contribution of genotypes to the
overall storage of real datasets (see the Genomics England example)
and the frequency that they are likely to be accessed, using 
Zstandard without BitShuffle 
would seem like a good tradeoff in many cases.
This ability to easily tune compression performance
and decoding speed on a field-by-field basis is a major strong
point of Zarr. The \texttt{vcf2zarr} utility also provides
functionality to aid with such storage schema tuning.

@@

\subsection{Zarr and block-based compression}
% Zarr is a simple layer on top of best-in-class, modern components
In the interest of completeness it is useful to provide a high-level overview
of Zarr and the technologies that it depends upon. Zarr is a specialised format
for storing large-scale $n$-dimensional data (arrays). Arrays
are split into chunks, which are compressed and stored separately. Chunks are 
addressed by their indexes along the dimensions of the array, and the 
compressed data associated with this key. Chunks can
be stored in individual files (as we do here), but a wide array of different
storage backends are supported including cloud object stores 
and NoSQL databases;
in principle, Zarr can store data in any key-value store.
Metadata describing the array and its properties is then stored 
in JSON format along with the chunks. 
The simplicity and transparency
of this design has substantial advantages over technologies
such as HDF~\citep{folk2011overview} and 
NetCDF~\citep{rew1990netcdf} which are based on complex
layouts of multidimensional data within a single file,
and cannot be accessed in practice without the corresponding library.
(See~\cite{abernathey2021cloud} for further discussion of the benefits
of Zarr over these monolithic file-oriented formats.)
In contrast, there are numerous implementations 
of the Zarr specification, ranging 
from the mature Zarr-Python~\citep{zarrpython}
and TensorStore~\citep{tensorstore} implementations
to more experimental extensions to packages like
GDAL~\citep{gdal_zarr},
NetCDF~\citep{netcfd_c},
N5~\citep{n5zarr}
and xtensor~\citep{xtensor_zarr}
as well as  standalone libraries for 
C++~\citep{ruan2024image}, JavaScript~\cite{zarrjs},
Julia~\cite{zarrjl}, Rust~\citep{zarrs}
and R~\cite{pizzarr}.

@@

>>>>

The data shown in Fig~\ref{fig-whole-matrix-compute} for Zarr and Savvy
is based on custom programs written using their respective APIs
to implement the \texttt{af-dist} operation. The Zarr program uses
the Zarr-Python package to iterate over the decoded chunks of the 
genotype matrix and classifies genotypes within a chunk using a 14 line Python
function, accelerated using the Numba JIT compiler~\cite{lam2015numba}.
The allele frequencies and genotype counts are then analysed to produce 
the final counts within the allele frequency bins with 9 lines of 
Python using NumPy~\cite{harris2020array} functions. Remarkably, this 
short and simple Python program is substantially faster than the 
equivalent compiled C using \texttt{htslib} APIs on BCF (6.9 hours
vs 20.6 hours for 1 million samples). 
% num_samples  num_sites      tool  user_time  sys_time     wall_time  total_time
%     1000000    7254858     savvy    8371.07      4.16   8377.718136  2.326453
%     1000000    7254858  bcftools   73939.32     46.09  74023.083678  20.551503
%     1000000    7254858      zarr   24709.64     29.32  24750.704171  6.871933
The fastest method is the 
C++ program written using the Savvy API. This would largely seem
to be due to Savvy's excellent genotype decoding performance
% zarr 1.2 GiB
% savvy 6.6 GiB
% zarr_nshf 2.7 GiB
% zarr_lz4_nshf 2.9 GiB
(up to 6.6GiB/s vs 1.2GiB/s for Zarr on this dataset;
Fig~\ref{fig-whole-matrix-decode}).
% sav / zarr ratio: [0.68777206 0.75870945 0.69865406 0.64644888 0.68131752
% 0.89460906]
%     sequence_length  num_samples  num_sites                tool         size
% 58         48129895      1000000    7254858                zarr    23.752762
% 50         48129895      1000000    7254858         10kx1k.zarr    22.068925
% 51         48129895      1000000    7254858                 bcf    51.749294
% 52         48129895      1000000    7254858             genozip    10.691384
% 53         48129895      1000000    7254858  lz4.noshuffle.zarr   214.868005
% 54         48129895      1000000    7254858      noshuffle.zarr    31.208149
% 55         48129895      1000000    7254858                 sav    21.249436
% 56         48129895      1000000    7254858                 tsk     1.802636
% 57         48129895      1000000    7254858                 vcf    81.375831
% 59         48129895      1000000    7254858            zarr.zip    24.240338
% 5          48129895      1000000    7254858                2bit  1689.153258
[TODO revise this text!!]
Turning off the BitShuffle filter for the Zarr dataset,
however, leads to a substantial increase in decoding speed
(2.9GiB/s) at the cost of a roughly 25\% increase in storage
space (29.9GiB up from 22.1GiB for 1 million samples; data not
shown). The LZ4 codec (again without BitShuffle) leads to slightly higher
decoding speed, but at the cost of much larger storage requirements 
(219GiB; data not shown).
Given the relatively small contribution of genotypes to the
overall storage of real datasets (see the Genomics England example)
and the frequency that they are likely to be accessed, using 
Zstandard without BitShuffle 
would seem like a good tradeoff in many cases.
This ability to easily tune compression performance
and decoding speed on a field-by-field basis is a major strong
point of Zarr. The \texttt{vcf2zarr} utility also provides
functionality to aid with such storage schema tuning.

A substantial advantage of Zarr is the availabilty of several 
independent implementations across different programming languages
(see Methods). Figure~\ref{fig-whole-matrix-compute-zarr-versions}
compares the CPU time required to perform the computations
of Figure~\ref{fig-whole-matrix-compute} 
for af-dist implementations using the Zarr-Python~\cite{zarrpython},
JZarr~\cite{jzarr},
and TensorStore~\citep{tensorstore} (C++ and Python) libraries.
The version using the Tensorstore C++ API is fastest, requiring 
about 20\% less time than Zarr-Python (which is used in other
benchmarks).

% num_samples  num_sites       tool  user_time  sys_time     wall_time
% storage  total_time
% 22      1000000    7254858       zarr   16709.55     38.14  16910.057917
% hdd    4.652136
% 20      1000000    7254858     ts_cpp   13233.50     92.81  13573.456810
% hdd    3.701753
% 21      1000000    7254858  zarr_java   43780.50    245.28  44134.462377
% hdd   12.229383
% 23      1000000    7254858      ts_py   18027.43    642.40  19163.174182
% hdd    5.186064
% Cpp / Zarr python [0.79197226]

@@

\subsection{Zarr and block-based compression}
% Zarr is a simple layer on top of best-in-class, modern components
In the interest of completeness it is useful to provide a high-level overview
of Zarr and the technologies that it depends upon. Zarr is a specialised format
for storing large-scale $n$-dimensional data (arrays). Arrays
are split into chunks, which are compressed and stored separately. Chunks are 
addressed by their indexes along the dimensions of the array, and the 
compressed data associated with this key. Chunks can
be stored in individual files (as we do here), but a wide array of different
storage backends are supported including cloud object stores 
and NoSQL databases;
in principle, Zarr can store data in any key-value store.
Metadata describing the array and its properties is then stored 
in JSON format along with the chunks. 
The simplicity and transparency
of this design has substantial advantages over technologies
such as HDF~\citep{folk2011overview} and 
NetCDF~\citep{rew1990netcdf} which are based on complex
layouts of multidimensional data within a single file,
and cannot be accessed in practice without the corresponding library.
(See~\cite{abernathey2021cloud} for further discussion of the benefits
of Zarr over these monolithic file-oriented formats.)
In contrast, there are numerous implementations 
of the Zarr specification, ranging 
from the mature Zarr-Python~\citep{zarrpython}
and TensorStore~\citep{tensorstore} implementations
to more experimental extensions to packages like
GDAL~\citep{gdal_zarr},
NetCDF~\citep{netcfd_c},
N5~\citep{n5zarr}
and xtensor~\citep{xtensor_zarr}
as well as  standalone libraries for 
C++~\citep{ruan2024image}, 
Java~\cite{jzarr},
JavaScript~\cite{zarrjs},
Julia~\cite{zarrjl}, Rust~\citep{zarrs}
and R~\cite{pizzarr}.

@@

\begin{figure}[h]
\includegraphics{figures/whole-matrix-compute-zarr-versions}
\caption{Whole-matrix computation performance using different Zarr 
implementations. 
Total CPU time required to run the af-dist calculation
the data in Figure~\ref{fig-data-storage} using different Zarr implementations.
Elapsed time is also reported (dotted line). 
These benchmarks were run on [TODO insert hardware details].
Note the difference between the
time reported here and in Fig~\ref{fig-whole-matrix-compute} for Zarr-Python
is due to different hardware platforms.
\label{fig-whole-matrix-compute-zarr-versions}}
\end{figure}
